A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-01-23 paper/paper.tex
paper: fix minor whitespace issues in Sigma types

\begin{enumerate}
  \item[(i)] UIP holds on the unit and bool type:
    \begin{align*}
      \textsf{eq\_refl}        & : \Pi x.\, x = x                                             \\
      \textsf{UIP\_refl\_unit} & : \Pi u: \unittype.\, \Pi x : u = u.\, x = \textsf{eq\_refl} \\
      \textsf{UIP\_refl\_bool} & : \Pi b: \booltype.\, \Pi x : b = b.\, x = \textsf{eq\_refl}
    \end{align*}
  \item[(ii)] UIP propagates to $\Sigma$-types.
  \item[(iii)] UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

@@

The latter correspondence can be expressed in Homotopy Type Theory, for $T: \Type$, as an equivalence between $\Sigma S: \Type\;.\;S \rightarrow T$ and $T \rightarrow \Type$, which can be proved using univalence. This correspondence also holds on \U, for $T: \U$:
\begin{equation*}
  \Sigma S: \U\;.\;S\rightarrow T ~\simeq~ T \rightarrow \U
\end{equation*}

\subsection{Example: indexed semi-cubical sets}
Let us now come back to semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.
\begin{equation*}
  \begin{tikzcd}
    X_0 : \U & X_1 : \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2 : \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar R}" description, shift right=2] \arrow[l, "\delta^{\kstar L}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying to the description above the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a : A, (B a \rightarrow C)$ and $(\Sigma a : A, B a) \rightarrow C$. Let us now illustrate how we build $\fullframe[n]$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Then, more generally, we let each $\fullframe[n]$ consist of $n$ layers, written $\layer[n][p]$ with $p < n$, that we stack in order, starting from the $\unittype$ type and writing $\framep[n][p]$ for the $p$ first layers of a $\fullframe[n]$, so that $\fullframe[n]$ is $\framep[n][n]$. For instance, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$, as shown in Fig.~\ref{fig:intuition} for $\nu = 2$ which also mentions how the type of $X_2$ is structured.

\begin{figure*}[!t]
  \begin{equation*}
    \begin{array}{ll}
      X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
      X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
      \begin{array}{c}
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
          \\ \times \\
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
        \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
      X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
          \begin{array}{c}
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\; \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L]}}_{\painting[1][1]}}_{\painting[1][0]}
            \\ \times \\
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\; \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R]}}_{\painting[1][1]}}_{\painting[1][0]}
          \end{array}
          \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
        \begin{array}{c}
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.L, \\
                               a.\tl.R.\hd.L
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][L]}}_{\painting[1][1]}
          \\ \times \\
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.R, \\
                               a.\tl.R.\hd.R
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][R]}}_{\painting[1][1]}
        \end{array}
        \right)}_{\layer[2][1]}}_{\framep[2][2]}
      \rightarrow \U                                                                                                                                \\
      \ldots
    \end{array}
  \end{equation*}
  \caption{Intuition for formal construction ($\nu = 2$)}\label{fig:intuition}
\end{figure*}

@@

We illustrated here the cubical case, that is $\nu = 2$, but, in general, a $\layer[n][p]$ is a product of $\nu$ $\painting[n - 1][p]$. A $\painting[n][0]$ is a $n$-dimensional object corresponding to a filled $\fullframe[n]$. More generally, a $\painting[n][p]$ is an $n$-dimensional object which has the form of a $\painting[n-p][0]$, thus of $(n-p)$-dimensional form, but shifted and living in dimensions $p$ to $n$. Such $\painting[n][p]$ fills a space framed by a partial $\framep[n][p]$ so that, together, they form a filled $\fullframe[n]$. For instance, in the picture, each of the two $\painting[2][0]$ of $\layer[3][0]$ is a filled blue square, each of the two $\painting[2][1]$ of $\layer[3][1]$ is the line shown in red, stretched into a partial square filling the partial frames made of respectively the left and right border of the blue square, and each of the two $\painting[2][2]$ of $\layer[3][2]$ is a point shown in green, stretched into a partial square filling the full frames made respectively of the upper and lower borders of the blue and red squares. A $\painting[n][p]$ adds to a $\framep[n][p]$ the layers needed to form a $\fullframe[n]$. They are added from dimension $n$ to dimension $p$ that is opposite to the order from $0$ to $p$ the $\framep[n][p]$ are built. A $\painting[n][p]$ complementing a given $\framep[n][p]$ fills the resulting $\fullframe[n]$. Thus, on the contrary of frames, it includes a filler, which lives in $X_n$ as shown below.
\begin{align*}
  \framep[n][p]   & \defeq \Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0]) \ldots) .\; \layer[n][p] \\
  \painting[n][p] & \defeq \Sigma l_p : \layer[n][p] . (\ldots (\Sigma l_n : \layer[n][n - 1] .\; X_n) \ldots)
\end{align*}

So far, we have not paid attention to the fact that we have a dependent type, shown as a $\Sigma$. Let us be more precise about this requirement. First, $\fullframe[n]$ depends on all $X_i$ up to $n - 1$. So, we need to package up $X_i$, for $i < n$, into a type, which we call $\X[][<n][]$. This allows us to give the type $\X[][<n][] \rightarrow \U$ to $\fullframe[n]$. Then, for $D: \X[][<n][]$, representing an initial prefix of $X_0, X_1, \ldots X_{n - 1}$, the indexed set $X_n$ has type $\fullframe[n](D) \rightarrow \U$. Thus, $\framep[n][p]$, $\layer[n][p]$ and $\painting[n][p]$ also depend on $D$. We can then refine the previous equation by showing the dependencies on $D$. In particular, $X_n$ is just $D.\tl$.
\begin{align*}
   & \framep[n][p](D) \defeq                                                                                     \\
   & \;\;\;\;\Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0](D)) \ldots) .\,\layer[n][p](D) \\
   & \painting[n][p](D) \defeq                                                                                   \\
   & \;\;\;\;\Sigma l_p : \layer[n][p](D).\, (\ldots (\Sigma l_n : \layer[n][n - 1](D) .\, D.\tl) \ldots)
\end{align*}

\begin{figure*}[!t]
  \begin{align*}
    \framep[n][p](D)      & \defeq & \Sigma d : (\ldots (\Sigma \unitpoint: \unittype.\, \layer[n][0](D)(\unitpoint)) \ldots).\, \layer[n][p](D)(d)                                \\
    \painting[n][p](D)(d) & \defeq & \Sigma l_p: \layer[n][p](D)(d).\, (\ldots (\Sigma l_n : \layer[n][n - 1](D)(d, l_p, \ldots, l_{n - 1}).\, D.\tl(d, l_p, \ldots, l_n)) \ldots) \\
                          &        & \text{where } (d, l_p, \ldots, l_q) \text{ abbreviates } ((\ldots(d, l_p), \ldots), l_q)
  \end{align*}
  \caption{Refinement of definition of $\framep$ and $\painting$\label{fig:refinement}}
\end{figure*}

@@

We now present the construction in \emph{extensional} type theory; i.e. in a type theory with the following reflection rule, where $=$ is propositional equality in some type and $\equiv$ is definitional equality~\cite{martinlof84}:
\begin{equation*}
  \seqr{}{\Gamma \vdash p : t = u}{\Gamma \vdash t \equiv u}
\end{equation*}

@@

The construction takes benefit of various provable equalities over proofs of equality being definitional by the reflection rule. This includes in particular the groupoid properties of equality. Notably, unicity of identity proofs holds in extensional type theory, so that any type is automatically an $\U$. Also, we left implicit in table~\ref{tab:coh} the use of the isomorphism between $u = v$ and $\Sigma (p:u.\hd = v.\hd). (u.\tl = v.\tl)$ for $u$ and $v$ in a $\Sigma$-type. In the same table, we also left implicit the use of the isomorphism between $f = g$ and $\Pi a:A.f(a) = g(a)$ for $f$ and $g$ in $\Pi a:A.\, B$, where it should be recalled that the right-to-left map, that is function extensionality, holds by default in extensional type theory.

>>>>

\begin{enumerate}
  \item[(i)] UIP holds on the unit and bool type:
    \begin{align*}
      \textsf{eq\_refl}        & : \Pi x.\, x = x                                            \\
      \textsf{UIP\_refl\_unit} & : \Pi u: \unittype.\, \Pi x: u = u.\, x = \textsf{eq\_refl} \\
      \textsf{UIP\_refl\_bool} & : \Pi b: \booltype.\, \Pi x: b = b.\, x = \textsf{eq\_refl}
    \end{align*}
  \item[(ii)] UIP propagates to $\Sigma$-types.
  \item[(iii)] UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

@@

The latter correspondence can be expressed in Homotopy Type Theory, for $T: \Type$, as an equivalence between $\Sigma S: \Type.\, S \rightarrow T$ and $T \rightarrow \Type$, which can be proved using univalence. This correspondence also holds on \U, for $T: \U$:
\begin{equation*}
  \Sigma S: \U.\, S\rightarrow T ~\simeq~ T \rightarrow \U
\end{equation*}

\subsection{Example: indexed semi-cubical sets}
Let us now come back to semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.
\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2: \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar R}" description, shift right=2] \arrow[l, "\delta^{\kstar L}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying to the description above the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a: A, (B a \rightarrow C)$ and $(\Sigma a: A, B a) \rightarrow C$. Let us now illustrate how we build $\fullframe[n]$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Then, more generally, we let each $\fullframe[n]$ consist of $n$ layers, written $\layer[n][p]$ with $p < n$, that we stack in order, starting from the $\unittype$ type and writing $\framep[n][p]$ for the $p$ first layers of a $\fullframe[n]$, so that $\fullframe[n]$ is $\framep[n][n]$. For instance, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$, as shown in Fig.~\ref{fig:intuition} for $\nu = 2$ which also mentions how the type of $X_2$ is structured.

\begin{figure*}[!t]
  \begin{equation*}
    \begin{array}{ll}
      X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
      X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
      \begin{array}{c}
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
          \\ \times \\
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
        \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
      X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
          \begin{array}{c}
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L]}}_{\painting[1][1]}}_{\painting[1][0]}
            \\ \times \\
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R]}}_{\painting[1][1]}}_{\painting[1][0]}
          \end{array}
          \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
        \begin{array}{c}
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.L, \\
                               a.\tl.R.\hd.L
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][L]}}_{\painting[1][1]}
          \\ \times \\
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.R, \\
                               a.\tl.R.\hd.R
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][R]}}_{\painting[1][1]}
        \end{array}
        \right)}_{\layer[2][1]}}_{\framep[2][2]}
      \rightarrow \U                                                                                                                                \\
      \ldots
    \end{array}
  \end{equation*}
  \caption{Intuition for formal construction ($\nu = 2$)}\label{fig:intuition}
\end{figure*}

@@

We illustrated here the cubical case, that is $\nu = 2$, but, in general, a $\layer[n][p]$ is a product of $\nu$ $\painting[n - 1][p]$. A $\painting[n][0]$ is a $n$-dimensional object corresponding to a filled $\fullframe[n]$. More generally, a $\painting[n][p]$ is an $n$-dimensional object which has the form of a $\painting[n-p][0]$, thus of $(n-p)$-dimensional form, but shifted and living in dimensions $p$ to $n$. Such $\painting[n][p]$ fills a space framed by a partial $\framep[n][p]$ so that, together, they form a filled $\fullframe[n]$. For instance, in the picture, each of the two $\painting[2][0]$ of $\layer[3][0]$ is a filled blue square, each of the two $\painting[2][1]$ of $\layer[3][1]$ is the line shown in red, stretched into a partial square filling the partial frames made of respectively the left and right border of the blue square, and each of the two $\painting[2][2]$ of $\layer[3][2]$ is a point shown in green, stretched into a partial square filling the full frames made respectively of the upper and lower borders of the blue and red squares. A $\painting[n][p]$ adds to a $\framep[n][p]$ the layers needed to form a $\fullframe[n]$. They are added from dimension $n$ to dimension $p$ that is opposite to the order from $0$ to $p$ the $\framep[n][p]$ are built. A $\painting[n][p]$ complementing a given $\framep[n][p]$ fills the resulting $\fullframe[n]$. Thus, on the contrary of frames, it includes a filler, which lives in $X_n$ as shown below.
\begin{align*}
  \framep[n][p]   & \defeq \Sigma a_n: (\ldots (\Sigma \unitpoint: \unittype .\, \layer[n][0]) \ldots) .\, \layer[n][p] \\
  \painting[n][p] & \defeq \Sigma l_p: \layer[n][p] . (\ldots (\Sigma l_n: \layer[n][n - 1] .\, X_n) \ldots)
\end{align*}

So far, we have not paid attention to the fact that we have a dependent type, shown as a $\Sigma$. Let us be more precise about this requirement. First, $\fullframe[n]$ depends on all $X_i$ up to $n - 1$. So, we need to package up $X_i$, for $i < n$, into a type, which we call $\X[][<n][]$. This allows us to give the type $\X[][<n][] \rightarrow \U$ to $\fullframe[n]$. Then, for $D: \X[][<n][]$, representing an initial prefix of $X_0, X_1, \ldots X_{n - 1}$, the indexed set $X_n$ has type $\fullframe[n](D) \rightarrow \U$. Thus, $\framep[n][p]$, $\layer[n][p]$ and $\painting[n][p]$ also depend on $D$. We can then refine the previous equation by showing the dependencies on $D$. In particular, $X_n$ is just $D.\tl$.
\begin{align*}
   & \framep[n][p](D) \defeq                                                                                  \\
   & \;\;\;\;\Sigma a_n: (\ldots (\Sigma \unitpoint: \unittype.\, \layer[n][0](D)) \ldots) .\,\layer[n][p](D) \\
   & \painting[n][p](D) \defeq                                                                                \\
   & \;\;\;\;\Sigma l_p: \layer[n][p](D).\, (\ldots (\Sigma l_n: \layer[n][n - 1](D) .\, D.\tl) \ldots)
\end{align*}

\begin{figure*}[!t]
  \begin{align*}
    \framep[n][p](D)      & \defeq & \Sigma d: (\ldots (\Sigma \unitpoint: \unittype.\, \layer[n][0](D)(\unitpoint)) \ldots).\, \layer[n][p](D)(d)                                \\
    \painting[n][p](D)(d) & \defeq & \Sigma l_p: \layer[n][p](D)(d).\, (\ldots (\Sigma l_n: \layer[n][n - 1](D)(d, l_p, \ldots, l_{n - 1}).\, D.\tl(d, l_p, \ldots, l_n)) \ldots) \\
                          &        & \text{where } (d, l_p, \ldots, l_q) \text{ abbreviates } ((\ldots(d, l_p), \ldots), l_q)
  \end{align*}
  \caption{Refinement of definition of $\framep$ and $\painting$\label{fig:refinement}}
\end{figure*}

@@

We now present the construction in \emph{extensional} type theory; i.e. in a type theory with the following reflection rule, where $=$ is propositional equality in some type and $\equiv$ is definitional equality~\cite{martinlof84}:
\begin{equation*}
  \seqr{}{\Gamma \vdash p: t = u}{\Gamma \vdash t \equiv u}
\end{equation*}

@@

The construction takes benefit of various provable equalities over proofs of equality being definitional by the reflection rule. This includes in particular the groupoid properties of equality. Notably, unicity of identity proofs holds in extensional type theory, so that any type is automatically an $\U$. Also, we left implicit in table~\ref{tab:coh} the use of the isomorphism between $u = v$ and $\Sigma (p:u.\hd = v.\hd). (u.\tl = v.\tl)$ for $u$ and $v$ in a $\Sigma$-type. In the same table, we also left implicit the use of the isomorphism between $f = g$ and $\Pi a: A.\, f(a) = g(a)$ for $f$ and $g$ in $\Pi a: A.\, B$, where it should be recalled that the right-to-left map, that is function extensionality, holds by default in extensional type theory.
